---
title: "Macroevolutionary Modeling of Continuous Trait Evolution"
author: "Jay Heejae Jang"
output: github_document
---

```{r setup, include=FALSE}

required_packages <- c("plotrix", "phytools", "TreeSim", "animation", "diverge", "tidyverse")
new_packages <- required_packages[!(required_packages %in% installed.packages()[, "Package"])]
if (length(new_packages) > 0) install.packages(new_packages)

invisible(lapply(required_packages, library, character.only = TRUE))

# Set a clean global theme for ggplot2
theme_set(theme_minimal() + theme(panel.grid.minor = element_blank()))
```


## In a Brownian motion model, the variance parameter (σ²) represents the evolutionary rate at which trait variance accumulates through time.The simulations below track cumulative BM random walks from an ancestral mysticete body size (4 m) under two distinct variance regimes over 100 time steps.

```{r}

set.seed(123)

n_step <- 100
start_point <- 4

# Generate normal distributions for step changes
normal1 <- rnorm(n_step, mean = 0, sd = 4)
normal2 <- rnorm(n_step, mean = 0, sd = 1)

random_walk1 <- cumsum(c(start_point, normal1))
random_walk2 <- cumsum(c(start_point, normal2))

# Adjust inner margins to prevent squeezed labels
par(mfrow = c(2, 1), mar = c(4, 4, 2, 1)) 

plot(random_walk1, type = 'l', col = 'dodgerblue3', lwd = 2,
     xlab = "Time", ylab = "Phenotypic Value", 
     main = "High Variance Regime (\u03c3 = 4)")

plot(random_walk2, type = 'l', col = 'deeppink3', lwd = 2,
     xlab = "Time", ylab = "Phenotypic Value", 
     main = "Low Variance Regime (\u03c3 = 1)")

par(mfrow = c(1, 1))
```

## To model clade-wide diversification, we simulate 76 independent evolutionary trajectories from a common ancestral state.

```{r}
set.seed(456)
n_species <- 76
n_steps <- 100
ancestral_size <- 4

# Generate matrix of evolutionary pathways
sim_clade <- function(sd_val) {
  replicate(n_species, cumsum(c(ancestral_size, rnorm(n_steps, 0, sd = sd_val)))) %>%
    as.data.frame() %>%
    mutate(Time = 0:n_steps) %>%
    pivot_longer(-Time, names_to = "Lineage", values_to = "Trait_Value")
}

clade_high <- sim_clade(sd_val = 4) %>% mutate(Regime = "High Rate (Var = 16)")
clade_low  <- sim_clade(sd_val = 1) %>% mutate(Regime = "Low Rate (Var = 1)")
clade_combined <- bind_rows(clade_high, clade_low)

# Plot clade disparity
ggplot(clade_combined, aes(x = Time, y = Trait_Value, group = Lineage)) +
  geom_line(alpha = 0.15, color = "midnightblue") +
  facet_wrap(~Regime) +
  labs(title = "Simulated Clade Disparity (76 Independent Lineages)",
       x = "Time Steps", y = "Body Length (m)")
```

## Evolution occurs along branching phylogenetic trees. Here, the BM process is projected onto simulated birth-death trees for pan-group (20 Myr) and crown-group (10 Myr) mysticetes. We map empirical data for modern giant genera alongside the simulated trait space to visualize how their observed body sizes compare with BM simulations.

```{r}
set.seed(6)

# Simulate phylogenetic trees
crown_tree <- sim.bd.age(age = 10, numbsim = 1, lambda = 0.2, mu = 0.05)[[1]]
pan_tree   <- sim.bd.age(age = 20, numbsim = 1, lambda = 0.2, mu = 0.05)[[1]]

# Simulate traits (Body Length: Var=1, Ancestor=4 | Tail Length: Var=0.5, Ancestor=2)
phylo_data <- bind_rows(
  tibble(
    Group = "Pan-group (Simulated)",
    Body_Length = fastBM(pan_tree, sig2 = 1, a = 4),
    Tail_Length = fastBM(pan_tree, sig2 = 0.5, a = 2)
  ),
  tibble(
    Group = "Crown-group (Simulated)",
    Body_Length = fastBM(crown_tree, sig2 = 1, a = 4),
    Tail_Length = fastBM(crown_tree, sig2 = 0.5, a = 2)
  )
)

# Empirical body and tail length values based on Slater et al. (2017).
empirical_data <- tibble(
  Group = c("Eubalaena (Empirical)", "Balaenoptera (Empirical)"),
  Body_Length = c(16, 24),    
  Tail_Length = c(3, 5))

all_traits <- bind_rows(phylo_data, empirical_data)

ggplot(all_traits, aes(x = Body_Length, y = Tail_Length, color = Group, shape = Group, size = Group)) +
  geom_point(alpha = 0.8) +
  scale_color_manual(values = c("turquoise", "violet", "pink", "blue")) +
  scale_shape_manual(values = c(18, 18, 17, 16)) +
  scale_size_manual(values = c(5, 5, 2, 2)) +
  labs(title = "Phylogenetic Trait Space: Simulated vs. Empirical",
       x = "Body Length (m)", y = "Tail Length (m)") +
  theme(legend.position = "bottom", legend.title = element_blank())
```


## Below, 50 independent lineages are simulated evolving toward an adaptive optimum of 10 over 25 time steps, contrasting strong vs. weak selective pull using Ornstein-Uhlenbeck (OU) Models

```{r}

set.seed(999)
N <- 50 
steps <- 25 
theta <- 10     # Adaptive optimum
sig2 <- 1       # Rate of evolution

par(mfrow = c(1, 2))

# Strong Selection
rw_ou_strong <- random_walks(model="OU_null", TIME = steps, steps = steps, nsim = N, 
                            sig2 = sig2, alpha = 1, theta = theta,
                            plot=TRUE, col=rep("darkslategrey", N), ylim=c(0,20), labels=FALSE)
axis(1, at=seq(0, 25, 5), labels=seq(0, 25, 5))            
title(xlab="Time Steps", ylab="Trait Value", main = 'Strong Optimum (\u03b1 = 1.0)') 
abline(h = theta, col = "red", lty = 2, lwd = 2)

# Weak Selection
rw_ou_weak <- random_walks(model="OU_null", TIME = steps, steps = steps, nsim = N, 
                          sig2 = sig2, alpha = 0.1, theta = theta,
                          plot=TRUE, col=rep("darkslategrey", N), ylim=c(0,20), labels=FALSE)
axis(1, at=seq(0, 25, 5), labels=seq(0, 25, 5))              
title(xlab="Time Steps", ylab="Trait Value", main = 'Weak Optimum (\u03b1 = 0.1)') 
abline(h = theta, col = "red", lty = 2, lwd = 2)

par(mfrow = c(1, 1))



```

## References

Slater, G. J., Goldbogen, J. A., & Pyenson, N. D. (2017). Independent evolution of baleen whale gigantism linked to Plio-Pleistocene ocean dynamics. *Proceedings of the Royal Society B: Biological Sciences, 284*(1855), 20170546. https://doi.org/10.1098/rspb.2017.0546
